---
title: "covariate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{covariate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cityClimateHealth)
```

# Types of covariates

covariates that occur at the `geo_unit` level

covariates that are at the exposure level

## Ways to add covariates to exposure 

(1) you add them after either the `exposure_matrix` object is created as:

    a. the special case of `is_us_holiday` which has its own function
    b. a single variable
    c. as another crossbasis variable

### method 1a: included in exposure matrix creation

```{r exposure}
library(dlnm)
library(gnm)
library(ggplot2)
library(data.table)

ex2 <- subset(ma_exposure, TOWN20 %in% c('BOSTON', 'CHELSEA'))

# create exposure matrix
exposure_columns <- list(
  "date" = "date",
  "exposure" = "tmax_C",
  "geo_unit" = "TOWN20",
  "geo_unit_grp" = "COUNTY20"
)

## basic
exposure_mat <- 
  make_exposure_matrix(
    ex2,
    exposure_columns, 
    time_subset = list(month = 5:9)
  )

head(exposure_mat)

```

### method 1a

```{r metho1a}

set.seed(123)
rr <- sample(1:nrow(ex2), 0.01 * nrow(ex2))
ex2$randCov <- 0
ex2$randCov[rr] <- 1

covariate_columns <- list(
  "date" = "date",
  "exposure" = "randCov",
  "geo_unit" = "TOWN20",
  "geo_unit_grp" = "COUNTY20"
)

# for a single variable you can do just maxlag = 0
# and we are doing exposure is factor = T
covariate_mat <- 
  make_exposure_matrix(
    ex2,
    covariate_columns, 
    maxlag = 2,
    exposure_is_factor = T,
    time_subset = list(month = 5:9)
  )

head(covariate_mat)
table(covariate_mat$randCov)
```

```{r join1}

## ok now join
exposure_mat <- join_exposure_covariate(exposure_mat, covariate_mat)
head(exposure_mat)
attributes(exposure_mat)$column_mapping
```

Special case of adding US holidays
```{r holidays}

# and add holiday
exposure_mat <- add_US_holiday(exposure_mat)
head(exposure_mat)
attributes(exposure_mat)$column_mapping
```


```{r ityouce} 
##
outcome_columns <- list(
  "date" = "date",
  "outcome" = "daily_deaths",
  "geo_unit" = "TOWN20",
  "geo_unit_grp" = "COUNTY20"
  
)

boston_deaths <- subset(ma_deaths, TOWN20 %in% c('BOSTON', 'CHELSEA'))

## basic
outcomes_tbl <- make_outcome_table(
  boston_deaths,  
  outcome_columns,
  time_subset = list(month = 5:9)
)
head(outcomes_tbl)

```



```{r run model2}
m1 <- condPois_1stage(exposure_mat,
                      outcomes_tbl, 
                      multi_zone = T,
                      global_cen = 25)

plot(m1)


```

# method 2b: included as a single variable

to do this just change maxlag = 0 in the step above

```{r sodfj12}

# for a single variable you can do just maxlag = 0
# and we are doing exposure is factor = T
covariate_mat <- 
  make_exposure_matrix(
    ex2,
    covariate_columns, 
    maxlag = 0,
    exposure_is_factor = T,
    time_subset = list(month = 5:9)
  )


```
