---
title: "Installing cityClimateHealth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing cityClimateHealth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To access "cityClimateHealth", users must first install CmdStanR, a command-line interface for Stan in R that translates statistical models into R code to execute Bayesian inference methods. 

To avoid memory limits or conflicts from other packages, it is best to run a new R session, or restart the current R session.

```{r, eval = FALSE}
install.packages('cmdstanr', repos = c('https://stan-dev.r-universe.dev', getOption('repos')))
```

After you've install `cmdstanr` try the `install_cmdstan()` function to download STAN.

If you're on MAC, you may need to install xcode
```{bash, eval = FALSE}
xcode-select --install
```

Now, users must install the "remotes" package in R.

```{r, eval = FALSE}
install.packages('remotes')
```

When these downloads are successfully completed, users can then install the cityClimateHealth package:

```{r, eval = FALSE}
library(remotes)
remotes::install_github("cmilando/cityClimateHealth")
```

