---
title: "Test of population sizes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Test of population sizes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cityClimateHealth)
```

This vignette 

Lets first make a simple function to estimate the attributable fraction, assuming a the following relationship between both exposure magnitude and exposure timing.

AF = rescale(z from 0 to 0.3)

where 

z is a function applied to the last three days of temperature

```{r z}
t0 = 25
z = function(t3, t2, t1) {
  ifelse(t3 > t0, 0.1 * (t3 - t0), 0) + 
  ifelse(t2 > t0, 0.2 * (t2 - t0), 0) + 
  ifelse(t1 > t0, 0.4 * (t1 - t0), 0)
}
```

Lets test that this works, so simulate some periodic exposure data:

```{r dlnmTest}
set.seed(123)
n = 1100
x = 1:n
aa = 2
bb  = 0.2
cc = 100 
vv = 23.25
ytrue = aa * sin(bb / (2*pi)* (x  + cc)) + vv
y = ytrue + rnorm(n, sd = 0.5)
plot(x, ytrue, 'l')
points(x, y, col = 'red')
abline(h = 25, col = 'brown')
```

And now get the underlying AF that are due to excess temperature
```{r}
AF_vec <- vector("numeric", n)
AF_vec[1] <- AF_vec[2] <- 0
for(i in 3:n) {
  local_z <- z(y[i], y[i-1], y[i-2])
  AF_vec[i] <- local_z
}

AF_vec <- scales::rescale(AF_vec, to = c(0, 0.3))
summary(AF_vec)
plot(x, AF_vec, 'l')
```

And now simulate baseline deaths -- this doesn't need to be related to the exposure beecause all we are tracking is excess mortality.
```{r}
baseline_deaths = runif(n, 100, 120)
```

And update deaths based on the AF

```{r update_deaths}
excess_deaths = baseline_deaths * AF_vec
new_deaths = baseline_deaths + excess_deaths
new_deaths = as.integer(round(new_deaths))
plot(x, new_deaths, 'l')
```

So now get an estimate of the exposure response function here, using our package

```{r getExposure}
exp_df <- data.frame(x, temp = y)
exp_df$dt = lubridate::make_date(2020, 1, 1) + exp_df$x
exp_df$TOWN20 = 'A_town'
exp_df$COUNTY20 = 'A_county'

exp_df$x <- NULL # a bug to fix

exposure_columns <- list(
  "date" = "dt",
  "exposure" = "temp",
  "geo_unit" = "TOWN20",
  "geo_unit_grp" = "COUNTY20"
)

exp_mat <- make_exposure_matrix(exp_df, exposure_columns)
head(exp_mat)
```
And now make the outcome dataset
```{r getOutcome}
deaths_df <- data.frame(dt = exp_df$dt,
                         deaths = new_deaths,
                         TOWN20 = 'A_town', 
                         COUNTY20 = 'A_county')

outcome_columns <- list(
  "date" = "dt",
  "outcome" = "deaths",
  "geo_unit" = "TOWN20",
  "geo_unit_grp" = "COUNTY20"
)

sim_tbl <- make_outcome_table(deaths_df,  outcome_columns)
sim_tbl
```
```{r testModel}
m1 <- condPois_1stage(exposure_matrix = exp_mat, 
                  outcomes_tbl = sim_tbl)
plot(m1)
```

Looks good (an interesting to note that there is a bump up in the exposure response curve at the lower end when we know there is not relationship there ... more on that later !)

Now the question we want to answer is, if this relationship is different for a larger
place and we put the two together in the same (i) 1-stage model or (ii) 2-stage model, how do things change -- is the larger county weighted more. 

Lets wrap our entire outcome dataset creation pipeline into a single function, assuming that the two spatial regions experience the same temperature. 

```{r fullOutcomePipeline} 
get_sim_outcome <- function(exposure_ts, t0, z1, z2, z3, 
                            AFmax, baseline_outcomes) {
  
  # the z function
  z = function(t3, t2, t1) {
    ifelse(t3 > t0, z1 * (t3 - t0), 0) + 
    ifelse(t2 > t0, z2 * (t2 - t0), 0) + 
    ifelse(t1 > t0, z3 * (t1 - t0), 0)
  }
  
  # AF 
  n = length(exposure_ts)
  AF_vec <- vector("numeric", n)
  AF_vec[1] <- AF_vec[2] <- 0
  for(i in 3:n) {
    local_z <- z(exposure_ts[i], exposure_ts[i-1], exposure_ts[i-2])
    AF_vec[i] <- local_z
  }

  AF_vec <- scales::rescale(AF_vec, to = c(0, AFmax))
  
  # outcomes
  excess_outcomes = baseline_outcomes * AF_vec
  new_outcomes = baseline_outcomes + excess_outcomes
  
  # make a df
  return(as.integer(round(new_outcomes)))
}
```

Confirm that it works

```{r confirm}
sim_out1 <- get_sim_outcome(exposure_ts = y, t0 = 25, 
                            z1 = 0.1, z2 = 0.2, z3 = 0.4,
                AFmax = 0.3, baseline_outcomes = baseline_deaths)

cor(sim_out1, new_deaths)
```

Great, now lets make two for different populations

There are probably a few important edge cases:

* the smaller population has an uncertain and null relationship
